---
title: "Log-logistic Dose-response Curves"
author: 
- "Christian Rubjerg Hejstvig-Larsen (brf337)"
- "Dina Gyberg Jensen (vbz248)"
- "Michaela Lukacova (dns525)"
institute: "University of Copenhagen"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      ratio: 16:9
      countIncrementalSlides: true
      navigation:
        scroll: false
---
<style type="text/css">
.remark-slide-content {
    font-size: 18px;
        padding: 1em 4em 1em 4em;
    }
.remark-slide-content > h1 {
  font-size: 40px;
}
.remark-slide-scaler {
    overflow-y: auto;
    overflow-x: auto;
}
</style>
```{r, include=FALSE, eval = FALSE}
rmarkdown::render('presentation.rmd')
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(CSwR)
style_mono_accent(
 base_color = "#4f7942")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(microbenchmark)
library(bench)
library(tidyverse)
library(profvis)
library(bench)
library(knitr)
theme_set(theme_bw() + theme(text = element_text(size = 13)))
```
###Introduction
Objective is to use stochastic optimization to estimate the parameters of a log-logistic dose-response model using non-linear least squares estimation. That is obtaining the parameters $\alpha, \beta, \gamma, \rho$ that minimize the loss function:
$$L(X,(\alpha,\beta,\gamma,\rho))=\frac{1}{N}\sum_{i=1}^N\left( y_i - f(x_i| \alpha, \beta,\gamma,\rho)\right)^2$$
<br>
Where the response is given by:
$$Y_i = f(x_i| \alpha, \beta,\gamma,\rho) + \varepsilon_i \hskip5mm \text{with}\hskip2mm \varepsilon_i \stackrel{iid}{\sim} N(0,\omega^2)$$
And the log-logistic dose-response model is given by:
$$f(x_i| \alpha, \beta,\gamma,\rho) = \gamma + \frac{\rho - \gamma}{1 + \exp(\beta \log(x_i) - \alpha)}$$

---
###Stochastic Gradient Descent
Start out implementing a standard version of the stochastic gradient descent algorithm (SGD). The gradient in a single point is given by
$$-2\cdot\nabla f(x_i| \alpha, \beta,\gamma,\rho)\cdot\left(y_i - f(x_i| \alpha, \beta,\gamma,\rho)\right)  $$

<br>
Where $\nabla f(x_i| \alpha, \beta,\gamma,\rho)$ is given by
$$\nabla f(x_i| \alpha, \beta,\gamma,\rho) = \begin{pmatrix}
  \frac{\rho-\gamma}{(1+\exp(\beta\log(x_i)-\alpha))^2}\cdot\exp(\beta\log(x_i)-\alpha)\\
  \frac{\rho-\gamma}{(1+\exp(\beta\log(x_i)-\alpha))^2}\cdot\exp(\beta\log(x_i)-\alpha)\cdot\log(x_i)\\
  1-\frac{1}{1+\exp(\beta\log(x_i)-\alpha)} \\
  \frac{1}{1+\exp(\beta\log(x_i)-\alpha)}
\end{pmatrix}$$
So the update scheme becomes:
$$\theta_{t+1} = \theta_t +2 \gamma_t\cdot\nabla f(x_i| \alpha, \beta,\gamma,\rho)\cdot\left(y_i - f(x_i| \alpha, \beta,\gamma,\rho)\right)$$
Where $\theta_t = (\alpha_t, \beta_t, \gamma_t, \rho_t)$ and $\gamma_t$ is the learning rate at time $t$.
---
###Implementation 
The gradient function and $f$ are implemented seperately
```{r}

sgd <- function(
    par,
    grad, # Function of parameter and observation index
    gamma, # Decay schedule or a fixed learning rate
    maxiter = 150, # Max epoch iterations
    sampler = sample, # How data is resampled. Default is a random permutation
    cb = NULL,
    epoch = vanilla,
    m = 1, # Batch size
    x,
    y,
    ...) {
  
  n <- length(x)
  gamma <- if (is.function(gamma)) gamma(1:maxiter) else rep(gamma, maxiter)
  
  for (k in 1:maxiter) {
    
    if (!is.null(cb)) cb()
    samp <- sampler(n)
    par <- epoch(par = par, samp = samp, gamma = gamma[k], 
                 grad = grad, n = n, x = x, y = y, m = m)
    
  }
  par
}

vanilla <- function(par, samp, gamma, grad, n, x, y, ...){
  for (j in 1:n) {
    i <- samp[j]
    par <- par - gamma * grad(par, x[i], y[i])
  }
  return(par)
}
```

---
###Sampling

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### Source code
source("~/comp_stat/Stochastic_Optimization_ML/SGD.R")
source("~/comp_stat/Stochastic_Optimization_ML/sampler.R")
source("~/comp_stat/Stochastic_Optimization_ML/GD.R")
```
 
 - Implemented an S3 object `parameters` and a method `sim` to simulate data from the log-logistic dose-response model.
 
 - 2 different types of sampling methods
 
  - `grid_sample`: samples $X$ from a fixed grid of points $(e, e^2,..., e^{15})$.
  
  - `gauss_sample`: samples $\log(X)$ from $\mathcal{N}(0, \omega^2)$. Note $\omega$ may not be too large.

- The method also allows for scaling the data, which is useful for optimization.

```{r, eval = FALSE, message=FALSE}
sim(parameters(1, 1, 1, 1), N = 100, grid = TRUE, scale = TRUE)
```
 - True parameters have been set to `c(2,5,1,2)` and the data is simulated with `N = 5000`.
 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(4027)
N <- 5000
param <- parameters(2,5,1,2)
data <- sim(param, N)
x <- data$x
y <- data$y
```

---
### Test of Algorithm 

We do a naive test of the algorithm, and run the algorithm with 3 different starting values: the true values $(2,5,1,2)$, $(1,1,1,1)$ and random values $(1.88, 0.31, 3.64, 2.29)$. Step size is $0.005$, maximal number of itereations $300$.  

```{r, echo = FALSE}
set.seed(4027)
sv1 <- param$par
sv2 <- c(1,1,1,1)
sv3 <- abs(rnorm(4, 2, 1.5))

sgd1 <- SGD(par0 = sv1, grad = grad, gamma = 0.005, x = x, y = y, true_par = param$par, maxit = 300)
sgd2 <- SGD(par0 = sv2, grad = grad, gamma = 0.005, x = x, y = y, true_par = param$par, maxit = 300)
sgd3 <- SGD(par0 = sv3, grad = grad, gamma = 0.005, x = x, y = y, true_par = param$par, maxit = 300)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
grid.arrange(plot(sgd1, 1) + ggtitle("True vals"),
             plot(sgd2, 1)+ ggtitle("c(1,1,1,1)"), 
             plot(sgd3, 1) + ggtitle("Random vals"),
             nrow = 1)
```

---
### Test of Algorithm 

We obtain quite different convergence schemes. 

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
plot(sgd1, 3) + 
  geom_line(aes(x = plot_data(sgd2)$.time, 
                y = plot_data(sgd2)$abs_dist_from_par), col = "orange") + 
  geom_line(aes(x = plot_data(sgd3)$.time, 
                y = plot_data(sgd3)$abs_dist_from_par), col = "red")

diff_start_vals_sgd <- tibble(
  "Par" = c("\u03B1 (2)", "\u03B2 (5)", "\u03B3 (1)", "\u03C1 (2)"),
  "True vals" = sgd1$est,
  "(1, 1, 1, 1)" = sgd2$est,
   "Random vals" = sgd3$est)

diff_start_vals_sgd %>% kable(caption = "Different starting values")
```

---
###Profiling the algorithm
```{r, echo = FALSE}
profvis({
f <- function(x, par){
  alpha <- par[1]
  beta <- par[2]
  gamma <- par[3]
  rho <- par[4]
  
  return(gamma + (rho - gamma) / (1 + exp(beta * log(x) - alpha)))
}



gradient <- function(par, i, x, y,...){
  alpha <- par[1]
  beta <- par[2]
  gamma <- par[3]
  rho <- par[4]
  
  x_i <- x[i]
  y_i <- y[i]
  
  expbetalogxalpha <- exp(beta * log(x_i) - alpha)
  
  identical_part <- - 2 * (y_i - f(x_i, par))
  
  grad_alpha <- mean(identical_part * (rho - gamma) * expbetalogxalpha / (1 + expbetalogxalpha)^2)
  grad_beta <- - mean(identical_part * (rho - gamma) * log(x[i]) * expbetalogxalpha / (1 + expbetalogxalpha)^2)
  grad_gamma <- mean(identical_part * (1 - 1 / (1 + expbetalogxalpha)))
  grad_rho <- mean(identical_part / (1 + expbetalogxalpha))
  
  return(c(grad_alpha, grad_beta, grad_gamma, grad_rho))
}

sgd <- function(
    par0,
    grad,
    N, # Sample size
    gamma, # Decay schedule or a fixed learning rate
    epoch = NULL,
    maxiter = 100, # Max epoch iterations
    sampler = sample, # How data is resampled. Default is a random permutation
    cb = NULL,
    ...) {
  
  if (is.function(gamma)) gamma <- gamma(1:maxiter)
  gamma <- rep_len(gamma, maxiter)
  
  par <- par0
  
  for (n in 1:maxiter) {
    if (!is.null(cb)) cb$tracer()
    
    samp <- sampler(N)
    
    if (is.null(epoch)){
       for (j in 1:N) {
         i <- samp[j]
        par <- par - gamma[n] * grad(par, i, ...)
       }
    } else {
      par <- epoch(par, samp, gamma[n], ...)
    }
  }
  par
}

sgd(c(3,3,3,3), grad = gradient, N = 1000, gamma = 0.01, x = x, y = y)
})
```
---
###Implementing the gradient function in RCPP
```{r}
library(Rcpp)

cppFunction('
NumericVector gradient_rcpp(NumericVector par, NumericVector indices, 
NumericVector x, NumericVector y) {
  // Extract parameters
  double alpha = par[0];
  double beta = par[1];
  double gamma = par[2];
  double rho = par[3];

  // Initialize gradients
  double grad_alpha = 0.0;
  double grad_beta = 0.0;
  double grad_gamma = 0.0;
  double grad_rho = 0.0;

  int n = indices.size(); // Number of indices

  // Loop over the indices
  for (int idx = 0; idx < n; ++idx) {
    int i = indices[idx] - 1;  // Convert from r to c++ indexing

    // Get individual data point
    double x_i = x[i];
    double y_i = y[i];

    // Calculating f(x_i, par)
    double f_x_i = gamma + (rho - gamma) / (1 + exp(beta * log(x_i) - alpha));

    // Exponential term
    double expbetalogxalpha = exp(beta * log(x_i) - alpha);

    // Identical part used in gradients
    double identical_part = -2 * (y_i - f_x_i);

    // Accumulate gradients for all indices
    grad_alpha += (identical_part * (rho - gamma) * expbetalogxalpha) 
    / pow(1 + expbetalogxalpha, 2);
    grad_beta += -(identical_part * (rho - gamma) * log(x_i) * expbetalogxalpha) 
    / pow(1 + expbetalogxalpha, 2);
    grad_gamma += identical_part * (1 - 1 / (1 + expbetalogxalpha));
    grad_rho += identical_part / (1 + expbetalogxalpha);
  }

  // Return the mean of accumulated gradients
  return NumericVector::create(grad_alpha / n, grad_beta / n, grad_gamma / n, grad_rho / n);
}
')

```

```{r}
gradient(c(3,3,3,3), c(1,2), x, y)
gradient_rcpp(c(3,3,3,3), c(1,2), x, y)
```

---
###Benchmarking
Benchmarking the RCPP implementation against the R implementation
```{r,echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
sim1 <- data[1:100,]

bench_results <- bench::mark(
    R = sgd(c(3,3,3,3), grad = gradient, N = 100, gamma = 0.01, x = sim1$x, y = sim1$y),
    Rcpp = sgd(c(3,3,3,3), grad = gradient_rcpp, N = 100, gamma = 0.01, x = sim1$x, y = sim1$y),
    iterations = 100,
    check = F
)
plot(bench_results)
```  
<br>
```{r,echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
knitr::kable(data.frame(expression = c('R', 'Rcpp'), bench_results[,2:9]))
```

---
```{r,echo=T, warning  = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
sim1 <- data[1:1000,]

bench_results <- bench::mark(
    R = sgd(c(3,3,3,3), grad = gradient, N = 1000, gamma = 0.01, x = sim1$x, y = sim1$y),
    Rcpp = sgd(c(3,3,3,3), grad = gradient_rcpp, N = 1000, gamma = 0.01, x = sim1$x, y = sim1$y),
    iterations = 10,
    check = F
)
plot(bench_results)

knitr::kable(data.frame(expression = c('R', 'Rcpp'), bench_results[,2:9]))
```

---
### Decay Schedule
*Der går noget galt her, bliver fikset i morgen:)*

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### Source code
source("~/comp_stat/Stochastic_Optimization_ML/SGD.R")
source("~/comp_stat/Stochastic_Optimization_ML/sampler.R")
source("~/comp_stat/Stochastic_Optimization_ML/GD.R")
```

We implement a flexible three-parameter power law family of decay schedules

$$
\gamma_n = \frac{\gamma_0 K}{K + n^{a}}
$$

We try a decay schedules with $a = 1$, $K= 0.4$ and $\gamma_0 = 1$. A different way to specify the decay schedule is to specify a desired learning rate $\gamma_1$ which should be reached at iteration $n_1$. These specifications then determine the parameter $K$. We specify two decay schedules both with $\gamma_0 = 1$, $n_1 = 100$ and $\gamma_1 = 0.01$. But one with $a = 1$ and one with $a = 2$. 

```{r, warning = FALSE, message = FALSE, echo = FALSE}
set.seed(4027)
decay1 <- decay_scheduler(gamma0 = 1, a = 1, K = 0.4)
decay2 <- decay_scheduler(gamma0 = 1, a = 1, n1 = 150, gamma1 = 0.01)
decay3 <- decay_scheduler(gamma0 = 1, a = 2, n1 = 150, gamma1 = 0.01)


sgd_decay1 <- SGD(par0 = sv1 + rep(0.1,4), grad = grad, gamma = decay1, x = x, y = y, true_par = param$par)
sgd_decay2 <- SGD(par0 = sv1+ rep(0.1,4), grad = grad, gamma = decay2, x = x, y = y, true_par = param$par)
sgd_decay3 <- SGD(par0 = sv1+ rep(0.1,4), grad = grad, gamma = decay3, x = x, y = y, true_par = param$par)
```

```{r, hide = TRUE, warning = FALSE, message = FALSE, echo = FALSE, fig.width=10, fig.height=4, fig.align='center'}
ggplot() +
  geom_line(aes(x = plot_data(sgd_decay1)$.time, 
                y = plot_data(sgd_decay1)$loss, color = "a = 1, K = 0.4"))+
  geom_line(aes(x = plot_data(sgd_decay2)$.time, 
                y = plot_data(sgd_decay2)$loss, color = "a = 1, gamma1 = 0.01")) +
  geom_line(aes(x = plot_data(sgd_decay3)$.time, 
                y = plot_data(sgd_decay3)$loss, color = "a = 2, gamma1 = 0.01")) +
  scale_y_log10() +
  labs(title = "Loss vs Time", x = "Time", y = "Loss")+
  scale_color_manual(values = c("a = 1, K = 0.4" = "red", 
                                "a = 1, gamma1 = 0.01" = "orange", 
                                "a = 2, gamma1 = 0.01" = "black"),
                         labels = c(
      expression(a == 1 ~ ", " ~ K == 0.4), 
      expression(a == 1 ~ ", " ~ gamma[1] == 0.01), 
      expression(a == 2 ~ ", " ~ gamma[1] == 0.01)))+
  labs(color = "Parameter values")
```

---
### Decay Schedule

```{r, hide = TRUE, warning = FALSE, message = FALSE, echo = FALSE, fig.width=10, fig.height=4, fig.align='center'}
ggplot() +
  geom_line(aes(x = plot_data(sgd_decay1)$.time, 
                y = plot_data(sgd_decay1)$abs_dist_from_par, color = "a = 1, K = 0.4"))+
  geom_line(aes(x = plot_data(sgd_decay2)$.time, 
                y = plot_data(sgd_decay2)$abs_dist_from_par, color = "a = 2, gamma1 = 0.1")) +
  geom_line(aes(x = plot_data(sgd_decay3)$.time, 
                y = plot_data(sgd_decay3)$abs_dist_from_par, color = "a = 2, gamma1 = 0.01")) +
  scale_y_log10() +
  labs(title = "Abs. dist. to par vs Time", x = "Time", y = "Abs. dist. to par")+
  scale_color_manual(values = c("a = 1, K = 0.4" = "red", 
                                "a = 2, gamma1 = 0.1" = "orange", 
                                "a = 2, gamma1 = 0.01" = "black"),
                         labels = c(
      expression(a == 1 ~ ", " ~ K == 0.4), 
      expression(a == 2 ~ ", " ~ gamma[1] == 0.1), 
      expression(a == 2 ~ ", " ~ gamma[1] == 0.01)
    ))+
  labs(color = "Parameter values")

```


---
### Gradient Descent

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### Source code
source("~/comp_stat/Stochastic_Optimization_ML/SGD.R")
source("~/comp_stat/Stochastic_Optimization_ML/sampler.R")
source("~/comp_stat/Stochastic_Optimization_ML/GD.R")
```

We have also implemented a gradient descent algorithm. We have created an equivalent class for this algorithm as well. As stopping criterion we use $|\theta_n - \theta_{n-1}| \leq \epsilon = 10^{-3}$. We use backtracking line search in the algorithm in order to find step length. We check that the algorithm works: 

```{r,  echo = FALSE, fig.width=10, fig.height=4, fig.align='center'}
gd0 <- GD(par = sv1 + rep(0.1,4), H = H, x = x, y = y, true_par = param$par)
gd1 <- GD(par = sv2, H = H, x = x, y = y, true_par = param$par)
gd2 <- GD(par = sv3, H = H, x = x, y = y, true_par = param$par)

ggplot() +
  geom_line(aes(x = plot_data(gd0)$.time, 
                y = plot_data(gd0)$loss, color = "a"))+
  geom_line(aes(x = plot_data(gd1)$.time, 
                y = plot_data(gd1)$loss, color = "b")) +
  geom_line(aes(x = plot_data(gd2)$.time, 
                y = plot_data(gd2)$loss, color = "c")) +
  scale_y_log10() +
  labs(title = "Loss vs Time", x = "Time", y = "Loss")+
  scale_color_manual(values = c("a" = "red", 
                                "b" = "orange", 
                                "c" = "black"),
                         labels = c("True vals", "c(1,1,1,1)", "Random vals"))+
  labs(color = "Starting values")


diff_start_vals_gd <- tibble(
  "Par" = c("\u03B1 (2)", "\u03B2 (5)", "\u03B3 (1)", "\u03C1 (2)"),
  "(2.1, 5.1, 1.1, 2.1)" = gd0$est,
  "(1, 1, 1, 1)" = gd1$est,
  "Random vals" = gd2$est)

diff_start_vals_gd %>% kable()
```


---
### Comparison

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### Source code
source("~/comp_stat/Stochastic_Optimization_ML/SGD.R")
source("~/comp_stat/Stochastic_Optimization_ML/sampler.R")
source("~/comp_stat/Stochastic_Optimization_ML/GD.R")
```

We compare the performance of the two algorithms. We use the same data as before and start the algorithms off in the points $(2.2, 5.2, 1.2, 2.2)$. 

```{r, echo = FALSE}
gd0 <- GD(par = param$par + rep(0.2,4), H = H, x = x, y = y, true_par = param$par)
sgd0 <- SGD(par0 = param$par + rep(0.2,4), grad = grad, x = x, y = y, true_par = param$par, 
            gamma = 0.01)
```

```{r, warning = FALSE, message = FALSE, echo = FALSE}
tibble(" " = c("\u03B1", "\u03B2", "\u03B3", "\u03C1"), 
       "True" = param$par, 
       "GD" = gd0$est, 
       "SGD" = sgd0$est) %>% kable() 
```

---
### Comparison

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.width=10, fig.height=4, fig.align='center'}
ggplot() +
  geom_line(aes(x = plot_data(gd0)$.time, 
                y = plot_data(gd0)$loss, color = "GD"))+
  geom_line(aes(x = plot_data(sgd0)$.time, 
                y = plot_data(sgd0)$loss, color = "SGD")) +
  scale_y_log10() +
  labs(title = "Loss vs Time", x = "Time", y = "Loss")+
  scale_color_manual(values = c("GD" = "red", "SGD" = "orange"))+
  labs(color = "Method")
ggplot() +
  geom_line(aes(x = plot_data(gd0)$.time, 
                y = plot_data(gd0)$abs_dist_from_par, color = "GD"))+
  geom_line(aes(x = plot_data(sgd0)$.time, 
                y = plot_data(sgd0)$abs_dist_from_par, color = "SGD")) +
  scale_y_log10() +
  labs(title = "Abs. dist to par", x = "Time", y = "Abs. dist to par")+
  scale_color_manual(values = c("GD" = "red", "SGD" = "orange"))+
  labs(color = "Method")
```




---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(gridExtra)
library(grid)
library(patchwork)

source("~/comp_stat/Christian/Assignment 4/Adaptive learning rate.R", local = knitr::knit_global())
source("~/comp_stat/Christian/Assignment 4/Stochastic gradient descent.R", local = knitr::knit_global())
source("~/comp_stat/Christian/Assignment 4/Sampling.R", local = knitr::knit_global())
```

###Mini-batch stochastic gradient descent
The idea is to calculate the gradient in a _batch_ of data points and update the parameters:

+ Sample $m$ indices, $I_n = \{i_1, ...,i_m \}$ from $\{1, ..., N\}$.
+ Compute $\rho_n = \frac{1}{m} \sum_{i \in I_n} \nabla L_\theta(x_i, y_i, \theta_n)$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \rho_n$

We sample a partition of $I_1 \cup I_2 \cup ... \cup I_{M} \subseteq \{1, ..., N\}$ for $M = \lfloor N/m \rfloor$.

```{r}
batch <- function(
    par,           # Parameter estimates
    samp,          # Sample of N indices
    gamma,         # Learning rate
    grad,          # Gradient function
    m = 50,        # Mini-batch size
    ...
){
  M <- floor(length(samp) / m) 
  for (j in 0:(M - 1)) {
    i <- samp[(j * m + 1):(j * m + m)]        # Sample m indices
    par <- par - gamma * grad(par, i, ...)    # Update parameter estimates
  }
  return(par)
}
```


```{r,echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
N <- 5000
omega <- 1 # Changing this significantly changes the output. High or small values makes the algorithm worse


# param <- parameters(2,5,1,2)
# data <- sim(param, N)
# x <- data$x
# y <- data$y

iterations <- 100
SGD_tracer <- tracer(c("par", "n"), Delta = 0)

m1 <- 20
m2 <- 50
m3 <- 250
m_values <- list(m1, m2, m3)

gamma1 <- 1e-1
gamma2 <- 1e-2
gamma3 <- 1e-5

rate1 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma1, n1 = iterations)
rate2 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma2, n1 = iterations)
rate3 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma3, n1 = iterations)
rate_values <- list(rate1, rate2, rate3)

true_par <- c(2, 5, 1, 2)
init_par <- sv3

# Function to create the plots dynamically
create_SGD_plot <- function(SGD_obj, plot_num) {
  plot(SGD_obj, plot_num) + 
    theme(axis.title.x = element_blank(), 
          axis.title.y = element_blank(), 
          plot.title = element_blank(),
          plot.margin = margin(10, 10, 20, 20))
}

```

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center'}
SGD_object_batchm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_batchm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_batchm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
batch_SGD_objects <- list(SGD_object_batchm1_1, SGD_object_batchm2_1, SGD_object_batchm3_1,
                          SGD_object_batchm1_2, SGD_object_batchm2_2, SGD_object_batchm3_2,
                          SGD_object_batchm1_3, SGD_object_batchm2_3, SGD_object_batchm3_3)

# Store all plots in a list
batch_SGD_plots <- lapply(batch_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
batch_grid <- grid.arrange(
  grobs = batch_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for batch algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
# Store all plots in a list
batch_SGD_plots <- lapply(batch_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
batch_grid <- grid.arrange(
  grobs = batch_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for batch algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

---

###Momentum
Version of batch gradient descent where we add _momemtum_ to the gradient through a convex combination of the current gradient and the previous gradient. Given $\theta_n$ and a batch $I_n$ with $|I_n| = m$ we

+ Compute $g_n = \frac{1}{m} \sum_{i \in I_n} \nabla_\theta L(y_i, x_i, \theta_n)$
+ Compute $\rho_n = \beta \rho_{n-1} + (1 - \beta) g_n$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \rho_n$

The value of $\beta$ determines the gradient memory and is a parameter that can be tuned. Default is set to $0.9$. Note that $\beta = 0$ corresponds to batch stochastic gradient descent.

```{r}
momentum <- function() {
  rho <- 0        # Initialize rho outside the inner function to keep track of the previous gradient
  function(
    par,          # Parameter values
    samp,         # Sample of N indices
    gamma,        # Learning rate
    grad,         # Gradient function
    m = 50,       # Mini-batch size
    beta = 0.9,   # Momentum memory
    ...
  ){
    M <- floor(length(samp) / m) 
    for (j in 0:(M - 1)) {
      i <- samp[(j * m + 1):(j * m + m)]
      rho <<- beta * rho + (1 - beta) * grad(par, i, ...)   # Using '<<-' assigns the value to rho in the enclosing environment
      par <- par - gamma * rho
    }
    par
  } 
}
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center'}
SGD_object_momentumm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_momentumm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_momentumm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
momentum_SGD_objects <- list(SGD_object_momentumm1_1, SGD_object_momentumm2_1, SGD_object_momentumm3_1,
                             SGD_object_momentumm1_2, SGD_object_momentumm2_2, SGD_object_momentumm3_2,
                             SGD_object_momentumm1_3, SGD_object_momentumm2_3, SGD_object_momentumm3_3)

# Store all plots in a list
momentum_SGD_plots <- lapply(momentum_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
momentum_grid <- grid.arrange(
  grobs = momentum_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for momentum algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
# Store all plots in a list
momentum_SGD_plots <- lapply(momentum_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
momentum_grid <- grid.arrange(
  grobs = momentum_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for momentum algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


---

###Adaptive learning rates
To mitigate tuning issues, we introduce the adam algorithm, an adaptive learning rate algorithm. The idea is to combine momemtum with a standardiziation of each coordinate direction of the descent direction. This is in practice done by dividing the learning rate by a running average of magnitude of previous gradients:
$$v_n = \beta_2 v_{n-1} + (1 - \beta_2) g_n \odot g_n$$
where we denote $\beta_2$ the forgetting factor. The complete algorithm is as follows:

+ Initialize $\theta_0$, $\rho_0 = 0$, $v_0 = 0$
+ Compute $g_n = \frac{1}{m} \sum_{i \in I_n} \nabla_\theta L(y_i, x_i, \theta_n)$
+ Compute $\rho_n = \beta_1 \rho_{n-1} + (1 - \beta_1) g_n$
+ Compute $v_n = \beta_2 v_{n-1} + (1 - \beta_2) g_n \odot g_n$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \frac{\rho_n}{\sqrt{v_n} + \epsilon}$

where we add $\epsilon$ to avoid division by zero (default is $\epsilon = 10^{-8}$). The interpretation of $\beta_1$ is the same as in the momentum algorithm.

---

```{r}
adam <- function() {
  rho <- v <- 0     # Initialize rho and v outside the inner function to keep track of the previous gradients
  function(
    par,            # Initial parameter values
    samp,           # Sample of N indices
    gamma,          # Learning rate
    grad,           # Gradient function
    m = 50,         # Mini-batch size
    beta1 = 0.9,    # First-moment memory
    beta2 = 0.9,    # Second-moment memory
    ...

  ){
    M <- floor(length(samp) / m) 
    for (j in 0:(M - 1)) {
      i <- samp[(j * m + 1):(j * m + m)]
      gr <- grad(par, i, ...)
      rho <<- beta1 * rho + (1 - beta1) * gr
      v <<- beta2 * v + (1 - beta2) * gr^2
      par <- par - gamma * (rho / (sqrt(v) + 1e-8))
    }
    par
  } 
}
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center'}
SGD_object_adamm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_adamm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_adamm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center', include = FALSE}
adam_SGD_objects <- list(SGD_object_adamm1_1, SGD_object_adamm2_1, SGD_object_adamm3_1,
                         SGD_object_adamm1_2, SGD_object_adamm2_2, SGD_object_adamm3_2,
                         SGD_object_adamm1_3, SGD_object_adamm2_3, SGD_object_adamm3_3)

# Store all plots in a list
adam_SGD_plots <- lapply(adam_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
adam_grid <- grid.arrange(
  grobs = adam_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for Adam algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```



```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center', include = F}
adam_SGD_objects <- list(SGD_object_adamm1_1, SGD_object_adamm2_1, SGD_object_adamm3_1,
                         SGD_object_adamm1_2, SGD_object_adamm2_2, SGD_object_adamm3_2,
                         SGD_object_adamm1_3, SGD_object_adamm2_3, SGD_object_adamm3_3)

# Store all plots in a list
adam_SGD_plots3 <- lapply(adam_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
adam_grid <- grid.arrange(
  grobs = adam_SGD_plots3,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for Adam algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


---

###Comparing convergence of the algorithms

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=11, fig.height=9, fig.align='center'}

p1 <- adam_SGD_plots[[1]] +
    geom_line(aes(x = plot_data(SGD_object_batchm1_1)$.time, 
                y = plot_data(SGD_object_batchm1_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_1)$.time, 
                y = plot_data(SGD_object_momentumm1_1)$loss), col = "red")

p2 <- adam_SGD_plots[[2]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_1)$.time, 
                y = plot_data(SGD_object_batchm2_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_1)$.time, 
                y = plot_data(SGD_object_momentumm2_1)$loss), col = "red")

p3 <- adam_SGD_plots[[3]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_1)$.time, 
                y = plot_data(SGD_object_batchm3_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_1)$.time, 
                y = plot_data(SGD_object_momentumm3_1)$loss), col = "red")

p4 <- adam_SGD_plots[[4]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_2)$.time, 
                y = plot_data(SGD_object_batchm1_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_2)$.time, 
                y = plot_data(SGD_object_momentumm1_2)$loss), col = "red")

p5 <- adam_SGD_plots[[5]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_2)$.time, 
                y = plot_data(SGD_object_batchm2_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_2)$.time, 
                y = plot_data(SGD_object_momentumm2_2)$loss), col = "red")

p6 <- adam_SGD_plots[[6]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_2)$.time, 
                y = plot_data(SGD_object_batchm3_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_2)$.time, 
                y = plot_data(SGD_object_momentumm3_2)$loss), col = "red")

p7 <- adam_SGD_plots[[7]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_3)$.time, 
                y = plot_data(SGD_object_batchm1_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_3)$.time, 
                y = plot_data(SGD_object_momentumm1_3)$loss), col = "red")

p8 <- adam_SGD_plots[[8]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_3)$.time, 
                y = plot_data(SGD_object_batchm2_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_3)$.time, 
                y = plot_data(SGD_object_momentumm2_3)$loss), col = "red")

p9 <- adam_SGD_plots[[9]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_3)$.time, 
                y = plot_data(SGD_object_batchm3_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_3)$.time, 
                y = plot_data(SGD_object_momentumm3_3)$loss), col = "red")


legend_grob <- grobTree(
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "black")),
  textGrob("Adam", x = unit(0.1, "npc"), y = 0.55, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "blue"), y = 0.4),
  textGrob("Mini-batch", x = unit(0.1, "npc"), y = 0.45, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "red"), y = 0.3),
  textGrob("Momentum", x = unit(0.1, "npc"), y = 0.35, just = "left")
)



momentum_grid <- grid.arrange(arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3),
                              legend_grob,
                              widths = c(9, 0.5, 0.5),
                              heights = c(9, 0.5, 0.5),
                              top = textGrob("Loss vs Time", gp = gpar(fontsize = 20, fontface = "bold")))


grid.text(paste0("m = ", m1), x = unit(0.15, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.45, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.8, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.55, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.25, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

---

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=11, fig.height=9, fig.align='center'}

p1 <- adam_SGD_plots3[[1]] +
    geom_line(aes(x = plot_data(SGD_object_batchm1_1)$.time, 
                y = plot_data(SGD_object_batchm1_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_1)$.time, 
                y = plot_data(SGD_object_momentumm1_1)$abs_dist_from_par), col = "red")

p2 <- adam_SGD_plots3[[2]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_1)$.time, 
                y = plot_data(SGD_object_batchm2_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_1)$.time, 
                y = plot_data(SGD_object_momentumm2_1)$abs_dist_from_par), col = "red")

p3 <- adam_SGD_plots3[[3]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_1)$.time, 
                y = plot_data(SGD_object_batchm3_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_1)$.time, 
                y = plot_data(SGD_object_momentumm3_1)$abs_dist_from_par), col = "red")

p4 <- adam_SGD_plots3[[4]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_2)$.time, 
                y = plot_data(SGD_object_batchm1_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_2)$.time, 
                y = plot_data(SGD_object_momentumm1_2)$abs_dist_from_par), col = "red")

p5 <- adam_SGD_plots3[[5]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_2)$.time, 
                y = plot_data(SGD_object_batchm2_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_2)$.time, 
                y = plot_data(SGD_object_momentumm2_2)$abs_dist_from_par), col = "red")

p6 <- adam_SGD_plots3[[6]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_2)$.time, 
                y = plot_data(SGD_object_batchm3_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_2)$.time, 
                y = plot_data(SGD_object_momentumm3_2)$abs_dist_from_par), col = "red")

p7 <- adam_SGD_plots3[[7]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_3)$.time, 
                y = plot_data(SGD_object_batchm1_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_3)$.time, 
                y = plot_data(SGD_object_momentumm1_3)$abs_dist_from_par), col = "red")

p8 <- adam_SGD_plots3[[8]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_3)$.time, 
                y = plot_data(SGD_object_batchm2_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_3)$.time, 
                y = plot_data(SGD_object_momentumm2_3)$abs_dist_from_par), col = "red")

p9 <- adam_SGD_plots3[[9]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_3)$.time, 
                y = plot_data(SGD_object_batchm3_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_3)$.time, 
                y = plot_data(SGD_object_momentumm3_3)$abs_dist_from_par), col = "red")





legend_grob <- grobTree(
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "black")),
  textGrob("Adam", x = unit(0.1, "npc"), y = 0.55, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "blue"), y = 0.4),
  textGrob("Mini-batch", x = unit(0.1, "npc"), y = 0.45, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "red"), y = 0.3),
  textGrob("Momentum", x = unit(0.1, "npc"), y = 0.35, just = "left")
)



momentum_grid <- grid.arrange(arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3),
                              legend_grob,
                              widths = c(9, 0.5, 0.5),
                              heights = c(9, 0.5, 0.5),
                              top = textGrob("Distance to true parameters vs Time", gp = gpar(fontsize = 20, fontface = "bold")))


grid.text(paste0("m = ", m1), x = unit(0.15, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.45, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.8, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.55, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.25, "npc"), rot = 90, gp = gpar(fontsize = 12))
```





---
### Exploiting Grid Structure 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### Source code
source("~/comp_stat/Stochastic_Optimization_ML/SGD.R")
source("~/comp_stat/Stochastic_Optimization_ML/sampler.R")
source("~/comp_stat/Stochastic_Optimization_ML/GD.R")
```

If we sample our $x$'s from a grid of $m$ points we can exploit the fact that we only have $m$ distinct $x$-values. We do not need to compute $f(x_i| \alpha, \beta,\gamma,\rho)$ for each $i$. We can compute the values for each distinct $x$-value and then use these values to compute the loss function. In the same way we can save computations when computing the derivative of $f$. We have implemented a `gd_grid` function that exploits this structure. 

```{r, warning = FALSE, message = FALSE}
gd_grid <- function(
    par,
    t0 = 1e-2,
    maxit = 1200,
    cb = NULL,
    epsilon = 1e-2,
    beta = 0.8,
    alpha = 0.1,
    x,
    y,
    ...) {
  
  x_vals <- unique(x)
  matches <- match(x, x_vals)
  n <- length(x)

  for (i in 1:maxit) {
    
    # Computing 
    fs <- f(par, x_vals)[matches]
    nabla_fs <- sapply(seq_along(x_vals), function(i) nabla_f(par, x_vals[i]))
    
    # Calculations of objective and gradient
    value <- sum((y - fs)^2) 
    gr <- - 2 / n * nabla_fs[,matches] %*% (y - fs)
    
    grad_norm <- sum(gr^2)
    
    # Callback
    if (!is.null(cb)) cb()
    
    t <- t0
    # Proposed descent step
    par_new <- par - t * gr
    new_fs <- f(par_new, x_vals)[matches]
    
    # Convergence criterion based on gradient norm
    if (all(abs(par_new - par) <= epsilon)) break
    
    # Backtracking line search
    while (sum((y - new_fs)^2) > value - alpha * t * grad_norm) {
      t <- beta * t
      par_new <- par - t * gr
      new_fs <- f(par_new, x_vals)[matches]
    }
    par <- par_new
  }
  
  if (i == maxit)  warning("Maximal number, ", maxit, ", of iterations reached")
  
  par
}

```

---
## Benchmarking the two GD versions

We sample data from a grid, we use the same true parametes $(2,5,1,2)$, and sample the $x$'s from the grid. 

```{r, warning = FALSE, message = FALSE, echo = FALSE}
data2 <- sim(param, N, grid = TRUE)
x2 <- data2$x
y2 <- data2$y
```

We check that the two algorithms return the same output:

```{r, warning = FALSE, message = FALSE, }
all(grad_desc(par = c(1,1,1,1), grad = grad_gd, H = H, x = x2, y = y2) -
  gd_grid(par = c(1,1,1,1), x = x2, y = y2)) < 10^(-8)
```

Comparing grid version with regular GD by benchmarking with 100 point

```{r, warning = FALSE, message = FALSE, echo = FALSE}
x_bench <- x2[1:100]
y_bench <- y2[1:100]

bench_results <- bench::mark(
    "GD" = grad_desc(par = sv1, grad = grad_gd, H = H, x = x2, y = y2),
    "Grid GD"= gd_grid(par = sv1, x = x2, y = y2),
    iterations = 100,
    check = F
)
plot(bench_results)
```

---

Comparing grid version with regular GD with 500 points. 

```{r, warning = FALSE, message = FALSE, echo = FALSE}
x_bench <- x2[1:1000]
y_bench <- y2[1:1000]

bench_results <- bench::mark(
    "GD" = grad_desc(par = sv1, grad = grad_gd, H = H, x = x2, y = y2),
    "Grid GD"= gd_grid(par = sv1, x = x2, y = y2),
    iterations = 100,
    check = F
)
plot(bench_results)
```
