---
title: "Log-logistic Dose-response Curves"
author: 
- "Christian Rubjerg Hejstvig-Larsen (brf337)"
- "Dina Gyberg Jensen (vbz248)"
institute: "University of Copenhagen"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      ratio: 16:9
      countIncrementalSlides: true
      navigation:
        scroll: false
---
<style type="text/css">
.remark-slide-content {
    font-size: 18px;
        padding: 1em 4em 1em 4em;
    }
.remark-slide-content > h1 {
  font-size: 40px;
}
.remark-slide-scaler {
    overflow-y: auto;
    overflow-x: auto;
}
</style>
```{r, include=FALSE, eval = FALSE}
rmarkdown::render('presentation.rmd')
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(CSwR)
style_mono_accent(
 base_color = "#4f7942")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(microbenchmark)
library(bench)
library(tidyverse)
library(profvis)
library(bench)
library(gridExtra)
library(grid)
library(patchwork)
theme_set(theme_bw() + theme(text = element_text(size = 13)))

source("/Users/christianhejstvig-larsen/Desktop/Studie/Master/First year/Block 1/Computational Statistics/comp_stat/Christian/Assignment 4/Adaptive learning rate.R", local = knitr::knit_global())
source("/Users/christianhejstvig-larsen/Desktop/Studie/Master/First year/Block 1/Computational Statistics/comp_stat/Christian/Assignment 4/Stochastic gradient descent.R", local = knitr::knit_global())
source("/Users/christianhejstvig-larsen/Desktop/Studie/Master/First year/Block 1/Computational Statistics/comp_stat/Christian/Assignment 4/Sampling.R", local = knitr::knit_global())
```

###Mini-batch stochastic gradient descent
The idea is to calculate the gradient in a _batch_ of data points and update the parameters:

+ Sample $m$ indices, $I_n = \{i_1, ...,i_m \}$ from $\{1, ..., N\}$.
+ Compute $\rho_n = \frac{1}{m} \sum_{i \in I_n} \nabla L_\theta(x_i, y_i, \theta_n)$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \rho_n$

We sample a partition of $I_1 \cup I_2 \cup ... \cup I_{M} \subseteq \{1, ..., N\}$ for $M = \lfloor N/m \rfloor$.

```{r}
batch <- function(
    par,           # Parameter estimates
    samp,          # Sample of N indices
    gamma,         # Learning rate
    grad,          # Gradient function
    m = 50,        # Mini-batch size
    ...
){
  M <- floor(length(samp) / m) 
  for (j in 0:(M - 1)) {
    i <- samp[(j * m + 1):(j * m + m)]        # Sample m indices
    par <- par - gamma * grad(par, i, ...)    # Update parameter estimates
  }
  return(par)
}
```


```{r,echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=4, fig.align='center'}
set.seed(4027)
N <- 5000
omega <- 1 # Changing this significantly changes the output. High or small values makes the algorithm worse


param <- parameters(2,5,1,2)
data <- sim(param, N)
x <- data$x
y <- data$y

iterations <- 100
SGD_tracer <- tracer(c("par", "n"), Delta = 0)

m1 <- 20
m2 <- 50
m3 <- 250
m_values <- list(m1, m2, m3)

gamma1 <- 1e-1
gamma2 <- 1e-2
gamma3 <- 1e-5

rate1 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma1, n1 = iterations)
rate2 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma2, n1 = iterations)
rate3 <- decay_scheduler(gamma0 = 1, a = 2, gamma1 = gamma3, n1 = iterations)
rate_values <- list(rate1, rate2, rate3)

true_par <- c(2, 5, 1, 2)
init_par <- abs(rnorm(4, mean = 2, sd = 1.5))

# Function to create the plots dynamically
create_SGD_plot <- function(SGD_obj, plot_num) {
  plot(SGD_obj, plot_num) + 
    theme(axis.title.x = element_blank(), 
          axis.title.y = element_blank(), 
          plot.title = element_blank(),
          plot.margin = margin(10, 10, 20, 20))
}

```

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center'}
SGD_object_batchm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_batchm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_batchm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_batchm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = batch, m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
batch_SGD_objects <- list(SGD_object_batchm1_1, SGD_object_batchm2_1, SGD_object_batchm3_1,
                          SGD_object_batchm1_2, SGD_object_batchm2_2, SGD_object_batchm3_2,
                          SGD_object_batchm1_3, SGD_object_batchm2_3, SGD_object_batchm3_3)

# Store all plots in a list
batch_SGD_plots <- lapply(batch_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
batch_grid <- grid.arrange(
  grobs = batch_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for batch algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
# Store all plots in a list
batch_SGD_plots <- lapply(batch_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
batch_grid <- grid.arrange(
  grobs = batch_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for batch algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

---

###Momentum
Version of batch gradient descent where we add _momemtum_ to the gradient through a convex combination of the current gradient and the previous gradient. Given $\theta_n$ and a batch $I_n$ with $|I_n| = m$ we

+ Compute $g_n = \frac{1}{m} \sum_{i \in I_n} \nabla_\theta L(y_i, x_i, \theta_n)$
+ Compute $\rho_n = \beta \rho_{n-1} + (1 - \beta) g_n$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \rho_n$

The value of $\beta$ determines the gradient memory and is a parameter that can be tuned. Default is set to $0.9$. Note that $\beta = 0$ corresponds to batch stochastic gradient descent.

```{r}
momentum <- function() {
  rho <- 0        # Initialize rho outside the inner function to keep track of the previous gradient
  function(
    par,          # Parameter values
    samp,         # Sample of N indices
    gamma,        # Learning rate
    grad,         # Gradient function
    m = 50,       # Mini-batch size
    beta = 0.9,   # Momentum memory
    ...
  ){
    M <- floor(length(samp) / m) 
    for (j in 0:(M - 1)) {
      i <- samp[(j * m + 1):(j * m + m)]
      rho <<- beta * rho + (1 - beta) * grad(par, i, ...)   # Using '<<-' assigns the value to rho in the enclosing environment
      par <- par - gamma * rho
    }
    par
  } 
}
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center'}
SGD_object_momentumm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_momentumm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_momentumm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_momentumm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = momentum(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
momentum_SGD_objects <- list(SGD_object_momentumm1_1, SGD_object_momentumm2_1, SGD_object_momentumm3_1,
                             SGD_object_momentumm1_2, SGD_object_momentumm2_2, SGD_object_momentumm3_2,
                             SGD_object_momentumm1_3, SGD_object_momentumm2_3, SGD_object_momentumm3_3)

# Store all plots in a list
momentum_SGD_plots <- lapply(momentum_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
momentum_grid <- grid.arrange(
  grobs = momentum_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for momentum algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=10, fig.height=8, fig.align='center', include = F}
# Store all plots in a list
momentum_SGD_plots <- lapply(momentum_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
momentum_grid <- grid.arrange(
  grobs = momentum_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for momentum algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


---

###Adaptive learning rates
To mitigate tuning issues, we introduce the adam algorithm, an adaptive learning rate algorithm. The idea is to combine momemtum with a standardiziation of each coordinate direction of the descent direction. This is in practice done by dividing the learning rate by a running average of magnitude of previous gradients:
$$v_n = \beta_2 v_{n-1} + (1 - \beta_2) g_n \odot g_n$$
where we denote $\beta_2$ the forgetting factor. The complete algorithm is as follows:

+ Initialize $\theta_0$, $\rho_0 = 0$, $v_0 = 0$
+ Compute $g_n = \frac{1}{m} \sum_{i \in I_n} \nabla_\theta L(y_i, x_i, \theta_n)$
+ Compute $\rho_n = \beta_1 \rho_{n-1} + (1 - \beta_1) g_n$
+ Compute $v_n = \beta_2 v_{n-1} + (1 - \beta_2) g_n \odot g_n$
+ Update $\theta_{n+1} = \theta_n - \gamma_n \frac{\rho_n}{\sqrt{v_n} + \epsilon}$

where we add $\epsilon$ to avoid division by zero (default is $\epsilon = 10^{-8}$). The interpretation of $\beta_1$ is the same as in the momentum algorithm.

---

```{r}
adam <- function() {
  rho <- v <- 0     # Initialize rho and v outside the inner function to keep track of the previous gradients
  function(
    par,            # Initial parameter values
    samp,           # Sample of N indices
    gamma,          # Learning rate
    grad,           # Gradient function
    m = 50,         # Mini-batch size
    beta1 = 0.9,    # First-moment memory
    beta2 = 0.9,    # Second-moment memory
    ...

  ){
    M <- floor(length(samp) / m) 
    for (j in 0:(M - 1)) {
      i <- samp[(j * m + 1):(j * m + m)]
      gr <- grad(par, i, ...)
      rho <<- beta1 * rho + (1 - beta1) * gr
      v <<- beta2 * v + (1 - beta2) * gr^2
      par <- par - gamma * (rho / (sqrt(v) + 1e-8))
    }
    par
  } 
}
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center'}
SGD_object_adamm1_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_1 <- SGD(par0 = init_par, grad = gradient, gamma = rate1,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_adamm1_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_2 <- SGD(par0 = init_par, grad = gradient, gamma = rate2,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)

SGD_object_adamm1_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m1, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm2_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m2, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
SGD_object_adamm3_3 <- SGD(par0 = init_par, grad = gradient, gamma = rate3,
                       N = N, epoch = adam(), m = m3, maxiter = iterations, sampler = sample,
                       cb = SGD_tracer, x = x, y = y,
                       true_par = true_par)
```


```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center', include = FALSE}
adam_SGD_objects <- list(SGD_object_adamm1_1, SGD_object_adamm2_1, SGD_object_adamm3_1,
                         SGD_object_adamm1_2, SGD_object_adamm2_2, SGD_object_adamm3_2,
                         SGD_object_adamm1_3, SGD_object_adamm2_3, SGD_object_adamm3_3)

# Store all plots in a list
adam_SGD_plots <- lapply(adam_SGD_objects, create_SGD_plot, plot_num = 1)

# Arrange the plots in a grid
adam_grid <- grid.arrange(
  grobs = adam_SGD_plots,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Loss vs Time for Adam algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```



```{r echo=F, warning  = FALSE, message = FALSE, fig.width=9, fig.height=8, fig.align='center', include = F}
adam_SGD_objects <- list(SGD_object_adamm1_1, SGD_object_adamm2_1, SGD_object_adamm3_1,
                         SGD_object_adamm1_2, SGD_object_adamm2_2, SGD_object_adamm3_2,
                         SGD_object_adamm1_3, SGD_object_adamm2_3, SGD_object_adamm3_3)

# Store all plots in a list
adam_SGD_plots3 <- lapply(adam_SGD_objects, create_SGD_plot, plot_num = 3)

# Arrange the plots in a grid
adam_grid <- grid.arrange(
  grobs = adam_SGD_plots3,
  ncol = 3, nrow = 3,
  widths = c(0.5, 0.5, 0.5),
  heights = c(0.5, 0.5, 0.5),
  top = textGrob("Distance to true parameters vs Time for Adam algorithm", gp = gpar(fontsize = 20, fontface = "bold"))
)

grid.text(paste0("m = ", m1), x = unit(0.17, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.5, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.83, "npc"), y = unit(0.01, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.5, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.15, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


---

###Comparing convergence of the algorithms

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=11, fig.height=9, fig.align='center'}

p1 <- adam_SGD_plots[[1]] +
    geom_line(aes(x = plot_data(SGD_object_batchm1_1)$.time, 
                y = plot_data(SGD_object_batchm1_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_1)$.time, 
                y = plot_data(SGD_object_momentumm1_1)$loss), col = "red")

p2 <- adam_SGD_plots[[2]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_1)$.time, 
                y = plot_data(SGD_object_batchm2_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_1)$.time, 
                y = plot_data(SGD_object_momentumm2_1)$loss), col = "red")

p3 <- adam_SGD_plots[[3]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_1)$.time, 
                y = plot_data(SGD_object_batchm3_1)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_1)$.time, 
                y = plot_data(SGD_object_momentumm3_1)$loss), col = "red")

p4 <- adam_SGD_plots[[4]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_2)$.time, 
                y = plot_data(SGD_object_batchm1_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_2)$.time, 
                y = plot_data(SGD_object_momentumm1_2)$loss), col = "red")

p5 <- adam_SGD_plots[[5]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_2)$.time, 
                y = plot_data(SGD_object_batchm2_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_2)$.time, 
                y = plot_data(SGD_object_momentumm2_2)$loss), col = "red")

p6 <- adam_SGD_plots[[6]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_2)$.time, 
                y = plot_data(SGD_object_batchm3_2)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_2)$.time, 
                y = plot_data(SGD_object_momentumm3_2)$loss), col = "red")

p7 <- adam_SGD_plots[[7]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_3)$.time, 
                y = plot_data(SGD_object_batchm1_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_3)$.time, 
                y = plot_data(SGD_object_momentumm1_3)$loss), col = "red")

p8 <- adam_SGD_plots[[8]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_3)$.time, 
                y = plot_data(SGD_object_batchm2_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_3)$.time, 
                y = plot_data(SGD_object_momentumm2_3)$loss), col = "red")

p9 <- adam_SGD_plots[[9]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_3)$.time, 
                y = plot_data(SGD_object_batchm3_3)$loss), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_3)$.time, 
                y = plot_data(SGD_object_momentumm3_3)$loss), col = "red")


legend_grob <- grobTree(
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "black")),
  textGrob("Adam", x = unit(0.1, "npc"), y = 0.55, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "blue"), y = 0.4),
  textGrob("Mini-batch", x = unit(0.1, "npc"), y = 0.45, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "red"), y = 0.3),
  textGrob("Momentum", x = unit(0.1, "npc"), y = 0.35, just = "left")
)



momentum_grid <- grid.arrange(arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3),
                              legend_grob,
                              widths = c(9, 0.5, 0.5),
                              heights = c(9, 0.5, 0.5),
                              top = textGrob("Loss vs Time", gp = gpar(fontsize = 20, fontface = "bold")))


grid.text(paste0("m = ", m1), x = unit(0.15, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.45, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.8, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.55, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.25, "npc"), rot = 90, gp = gpar(fontsize = 12))
```

---

```{r echo=F, warning  = FALSE, message = FALSE, fig.width=11, fig.height=9, fig.align='center'}

p1 <- adam_SGD_plots3[[1]] +
    geom_line(aes(x = plot_data(SGD_object_batchm1_1)$.time, 
                y = plot_data(SGD_object_batchm1_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_1)$.time, 
                y = plot_data(SGD_object_momentumm1_1)$abs_dist_from_par), col = "red")

p2 <- adam_SGD_plots3[[2]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_1)$.time, 
                y = plot_data(SGD_object_batchm2_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_1)$.time, 
                y = plot_data(SGD_object_momentumm2_1)$abs_dist_from_par), col = "red")

p3 <- adam_SGD_plots3[[3]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_1)$.time, 
                y = plot_data(SGD_object_batchm3_1)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_1)$.time, 
                y = plot_data(SGD_object_momentumm3_1)$abs_dist_from_par), col = "red")

p4 <- adam_SGD_plots3[[4]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_2)$.time, 
                y = plot_data(SGD_object_batchm1_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_2)$.time, 
                y = plot_data(SGD_object_momentumm1_2)$abs_dist_from_par), col = "red")

p5 <- adam_SGD_plots3[[5]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_2)$.time, 
                y = plot_data(SGD_object_batchm2_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_2)$.time, 
                y = plot_data(SGD_object_momentumm2_2)$abs_dist_from_par), col = "red")

p6 <- adam_SGD_plots3[[6]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_2)$.time, 
                y = plot_data(SGD_object_batchm3_2)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_2)$.time, 
                y = plot_data(SGD_object_momentumm3_2)$abs_dist_from_par), col = "red")

p7 <- adam_SGD_plots3[[7]] +
      geom_line(aes(x = plot_data(SGD_object_batchm1_3)$.time, 
                y = plot_data(SGD_object_batchm1_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm1_3)$.time, 
                y = plot_data(SGD_object_momentumm1_3)$abs_dist_from_par), col = "red")

p8 <- adam_SGD_plots3[[8]] +
      geom_line(aes(x = plot_data(SGD_object_batchm2_3)$.time, 
                y = plot_data(SGD_object_batchm2_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm2_3)$.time, 
                y = plot_data(SGD_object_momentumm2_3)$abs_dist_from_par), col = "red")

p9 <- adam_SGD_plots3[[9]] +
      geom_line(aes(x = plot_data(SGD_object_batchm3_3)$.time, 
                y = plot_data(SGD_object_batchm3_3)$abs_dist_from_par), col = "blue") +
  geom_line(aes(x = plot_data(SGD_object_momentumm3_3)$.time, 
                y = plot_data(SGD_object_momentumm3_3)$abs_dist_from_par), col = "red")





legend_grob <- grobTree(
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "black")),
  textGrob("Adam", x = unit(0.1, "npc"), y = 0.55, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "blue"), y = 0.4),
  textGrob("Mini-batch", x = unit(0.1, "npc"), y = 0.45, just = "left"),
  rectGrob(width = 0.5, height = 0.05, gp = gpar(fill = "red"), y = 0.3),
  textGrob("Momentum", x = unit(0.1, "npc"), y = 0.35, just = "left")
)



momentum_grid <- grid.arrange(arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3),
                              legend_grob,
                              widths = c(9, 0.5, 0.5),
                              heights = c(9, 0.5, 0.5),
                              top = textGrob("Distance to true parameters vs Time", gp = gpar(fontsize = 20, fontface = "bold")))


grid.text(paste0("m = ", m1), x = unit(0.15, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m2), x = unit(0.45, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))
grid.text(paste0("m = ", m3), x = unit(0.8, "npc"), y = unit(0.1, "npc"), gp = gpar(fontsize = 12))

# Add y-axis labels for gamma-values (next to the left column)
grid.text(bquote(gamma[1] == .(gamma1)), x = unit(0.01, "npc"), y = unit(0.85, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma2)), x = unit(0.01, "npc"), y = unit(0.55, "npc"), rot = 90, gp = gpar(fontsize = 12))
grid.text(bquote(gamma[1] == .(gamma3)), x = unit(0.01, "npc"), y = unit(0.25, "npc"), rot = 90, gp = gpar(fontsize = 12))
```


