---
title: 'Univariate Simulation A: Rejection Sampling'
author: "Michaela Lukacova"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  fig_height: 5
  fig_width: 5
  theme: flatly
  html_document: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r source, include=FALSE}
source("~/Desktop/Uni/5aar/CompStat/Assignments/Univariate simulation/gaussian_envelope.R")
```

# Rejection Sampling

*Find a Gaussian envelope of $f$ and implement rejection sampling from the distribution with density f using this envelope.*

*Implement the adaptive rejection sampling algorithm that uses a piecewise log-affine envelope and compare it with the one based on the Gaussian envelope.*

*You are welcome to invent other envelopes and try to optimize the choice of envelope. However, for the exam the focus should be on the implementations and their comparisons and not on theoretical derivations of envelopes.*

## Gaussian envelope

We define the function that is proportional to our target density.
$$
f(x) \propto \prod_{i=1}^{100} \exp \left( yx_i z_i - \exp(yx_i) \right) =  \exp \left( y\left(\sum_{i=1}^{100}x_i z_i\right) - \sum_{i=1}^{100}\exp(yx_i) \right) = f^*(y)
$$

Note that

$$
1 = \int f(x) dx = c \int f^*(x) dx \Leftrightarrow \frac{1}{\int f(x) dx} = c
$$
We can thus find the proportionality constant by integrating the function $f^*$.

```{r}
# Vectorize function to be able to plot and integrate
f_star_vec <- Vectorize(f_star)

c <- 1 / integrate(f_star_vec, lower = 0, upper = Inf, abs.tol = 10^(-60))$value
c
```
A plot of the density 

```{r}
ggplot() +
  geom_function(fun = f_star_vec) +  
  labs(x = expression(y),
       y = expression({f^{"*"}}(y)))
```

In order to do rejection sampling one needs an envelope. We wish to use a Gaussian envelope $g = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( - \frac{(y- \mu)^2}{2\sigma^2}\right)$. The Gaussian density has two parameters, the mean and the variance. The envelope should be as tight as possible, and the density looks quite symmetric, therefore as the optimal mean of the Gaussian envelope we use the max of $f^*$ function. 

```{r}
mu_opt <- optimize(function(y) -f_star(y), interval = c(0.2, 0.3))$minimum
mu_opt
```
There is still another parameter to determine - the variance. The optimal variance parameter can be determined by considering the acceptance rates associated to the different variances, and picking the variance that maximizes the acceptance rate. The acceptance rate $\alpha$ is proportional to $\alpha'$ (at least as long as $c$ and $d$ are fixed). $\alpha'$ should fulfill
$$f^*(y) \alpha' \leq g(y)$$
We determine $\alpha'$ by considering the ratio of the $g(y)$ and $f^*(y)$. We are interested in the convexity of the ratio and a potential minimum. These objectives will be the same for the log of the ratios.
$$
\log \left( \frac{g(y)}{f^*(y)}\right) = -\log(\sqrt{2\pi\sigma^2})-\frac{(y-\mu_{opt})^2}{2\sigma^2} - y\left(\sum_{i=1}^{100}x_i z_i\right) + \sum_{i=1}^{100}\exp(yx_i)
$$

The first derivative is
$$
\frac{d}{dy} \log \left( \frac{g(y)}{f^*(y)}\right) = -\frac{y - \mu_{opt}}{\sigma^2} -\left(\sum_{i=1}^{100}x_i z_i\right) + \sum_{i=1}^{100}x_i \exp(yx_i)
$$
The second derivative is 
$$
\frac{d^2}{dy^2} \log \left( \frac{g(y)}{f^*(y)}\right) = - \frac{1}{\sigma^2} + \sum_{i=1}^{100} x_i^2 \exp(yx_i)
$$
The second derivative is not always positive, and the function is thus not convex. It will therefore be difficult to determine the global minimum by optimization. Below we plot the log ratio for a given $\sigma = 0.05$.

```{r}
log_ratio_vec <- Vectorize(log_ratio)
plot_ratio <- function(y) log_ratio_vec(y, mu = mu_opt, sigma = 0.05)


ggplot() +
  geom_function(fun = plot_ratio1) +
  xlim(0.01,0.6)+  
  labs(x = expression(y),
       y = "log ratio")
```

We will, since this is a relatively small problem, resort to a simple way of finding the infimum, namely taking the minimum over a grid of points. This minimum will be our $\alpha'$. 
As mentioned, the optimal variance is the variance that maximizes the acceptance rate, that is

$$
\sigma_{opt} = \arg \max_{\sigma} \left( \inf_{y} \log \left( \frac{g(y)}{f^*(y)} \right) \right)
$$
We for a sequence of sigma values determine the minimum of the logged ratio and pick the sigma corresponding to the largest value. 
```{r}
sigma_vals <- seq(0.01, 0.4, length.out = 10^3)
y_vals <- seq(0.01, 1, length.out = 10^4)
alpha_prime <- numeric(10^4)

for(i in seq_along(sigma_vals)){
  log_ratio_vals <- log_ratio_vec(y_vals, mu = mu_opt, sigma = sigma_vals[i])
  alpha_prime[i] <- min(log_ratio_vals)
}

# Optimal sigma
max_i <- which(alpha_prime == max(alpha_prime))
sigma_opt <- sigma_vals[max_i]
sigma_opt

# alpha prime
alpha_p <- alpha_prime[max_i] %>% exp()
alpha_p

# Envelope
g <- function(y) 1 / (sqrt(2 * pi *sigma_opt^2)) * exp(-(y - mu_opt) ^ 2 / (2 * sigma_opt ^ 2))
```

```{r}
f_star_alpha <- function(y) f_star_vec(y) * alpha_p

ggplot() +
  geom_function(fun = f_star_alpha) + 
  geom_function(fun = g, color = "red", linetype = "dashed")+
  labs(x = expression(y),
       y = expression(density))


sum(g(seq(0,1, by = 0.001)) - f_star_vec(seq(0,1, by = 0.001) * alpha_p) <= 0)
```
We define a rejection sampling function that takes as input $n$ and returns a vector $y$ of accepted draws and the acceptance rate

```{r}
# generate
samp0 <- f_sample(10^5)

#Acceptance rate is
alpha_est <- samp0$Accept
```

Using this acceptance rate as an estimate for the true $\alpha$,  the proportionality constant between $f$ and $f^*$ can be calculated by
$$
c = \frac{\alpha' d }{\hat{\alpha}} =  \frac{\alpha'}{\sqrt{2\pi}\sigma_{opt}\hat{\alpha}}
$$

```{r}
c <- alpha_p / (sqrt(2*pi) * sigma_opt * alpha_est)
f_est <- function(y) f_star_vec(y) * c

# plot histogram
ggplot(data = tibble(x = seq(1,10^5), y = samp0$Sample))+
  geom_histogram(aes(x = y, y = ..density..), color = "white")+
  geom_function(fun = f_est, color = "steelblue")
```
Evaluating the $f^*$ and $g^*$ take up time. And calling the random generators in an sequential manner is costly. 
```{r}
profvis::profvis({
  f_sample(10000)
})
```

We improve by implementing a vectorized version

```{r}
vec_f_samp <- rng_wrapper(vec_f_random)
samp1 <- vec_f_samp(10^5)

#Acceptance rate is
samp1$Accept
alpha_est <- samp1$Accept
```

```{r}
# plot histogram
c <- alpha_p / (sqrt(2*pi) * sigma_opt * alpha_est)
f_est <- function(y) f_star_vec(y) * c

# plot histogram
ggplot(data = tibble(x = seq(1,10^5), y = samp0$Sample))+
  geom_histogram(aes(x = y, y = ..density..), color = "white")+
  geom_function(fun = f_est, color = "steelblue")
```
Looks about the same. Profiling
```{r}
profvis::profvis({
  vec_f_samp(10000)
})
```
Evaluating $f^*$ takes up a lot of time. 

```{r}
rng_bench <- bench::press(
  k = 10^(2:4),
  {
    bench::mark(
      "Naive" = f_sample(k, alpha_p, mu_opt, sigma_opt),
      "Vectorized" = vec_f_samp(k),
      check = FALSE
    )
  }
)

rng_bench %>% 
  mutate(expr = as.character(expression), median = as.numeric(median)) %>% 
  ggplot(aes(k, median, color = expr)) + geom_point() + scale_y_log10() +
  geom_line() + labs(y = "time (ms)")

autoplot(rng_bench)
```
The new version is quite a lot faster.

## Piecewise affine envelope
Since

$$
\frac{d^2}{dy^2}  \left( y\left(\sum_{i=1}^{100}x_i z_i\right) - \sum_{i=1}^{100}\exp(yx_i) \right) = - \sum_{i=1}^{100}x_i^2 \exp(yx_i) < 0
$$
Our target density is proportional to a log concave function. We choose the $y$'s as $y_1 = -2.5, y_2 = -2, ..., y_8 = 1$ and calculate the $a_i$'s according to CSR. 

```{r}
ys <- seq(0.1,0.4, by = 0.01)

pa_sample <- rng_wrapper(rf)
sample2 <- pa_sample(10^5, ys)

# Acceptance rate
sample2$Accept

# plot histogram
c <- alpha_p / (sqrt(2*pi) * sigma_opt * sample2$Accept)
f_est <- function(y) f_star_vec(y) * c

# plot histogram
ggplot(data = tibble(x = seq(1,10^5), y = samp0$Sample))+
  geom_histogram(aes(x = y, y = ..density..), color = "white")+
  geom_function(fun = f_est, color = "steelblue")
```

### To do:
- Find sande alpha
- Kig mere på den affine (profile, benchmark, gør den pænere, plot den osv.)
- Implementer funktioner i rcpp
- gør det hele pænere
- evt. lav f* til matrix
- skab objekter
- tjek noter for andre ting som skal ordnes




