---
title: "Rejection Sampling"
author: "Michaela Lukacova (dns525)"
institute: "University of Copenhagen"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: themer-new.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      ratio: 16:9
      countIncrementalSlides: true
      navigation:
        scroll: false
---
<style type="text/css">
.remark-slide-content {
    font-size: 18px;
        padding: 1em 4em 1em 4em;
    }
.remark-slide-content > h1 {
  font-size: 40px;
}
.remark-slide-scaler {
    overflow-y: auto;
    overflow-x: auto;
}
</style>

```{r, include=FALSE, eval = FALSE}
rmarkdown::render('presentation.rmd')
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
library(CSwR)
style_mono_accent(
 base_color = "#4f7942")
```

```{r packages, echo=FALSE, warning=FALSE, message=FALSE}
library(microbenchmark)
library(bench)
library(tidyverse)
library(profvis)
library(bench)
library(knitr)
library(testthat)
knitr::opts_chunk$set(fig.retina = 2)
theme_set(
  theme_bw(base_size = 18, base_family = 'Noto Sans') +  # Set base font size and family
    theme(
      text = element_text(size = 15),           # Adjust text size
      axis.line = element_line(size = 4),       # Set axis line thickness (use `size` not `linewidth`)
      panel.grid = element_line(size = 0.5)     # Set grid line thickness
    )
)
```

### Introduction and outline

- We are given a Poisson data set with 100 rows and $2$ variables $z \in \mathbb{N}$ and $x \in \mathbb{R}_+$

- We wish to sample from the distribution on $[0, \infty)$ with density function 
$$f(y) \propto \prod_{i=1}^{100} \exp \left( yx_i z_i - \exp(yx_i) \right) =  \exp \left( y\left(\sum_{i=1}^{100}x_i z_i\right) - \sum_{i=1}^{100}\exp(yx_i) \right) = f^*(y)$$
- We wish to do this by
  + Finding a Gaussian envelope of $f$  
  + Implement rejection sampling from the distribution with density $f$ using this envelope
  
- We will further implement the adaptive rejection sampling algorithm that uses a piecewise log-affine envelope

- And we will compare of the two methods

---
### The Gaussian Envelope

To do rejection sampling one needs an envelope. The envelope should satisfy



We wish to use a Gaussian envelope, which means that We start out by plotting the function $f^*$ that we are given.  

```{r, echo = FALSE}
ggplot() +
  geom_function(fun = f_star_vec) +  
  labs(x = expression(y),
       y = expression({f^{"*"}}(y)))
```

---
###


$$g(y) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( - \frac{(y- \mu)^2}{2\sigma^2}\right)$$
- The Gaussian density has two parameters, the mean $\mu$ and the variance $\sigma^2$ 
- The envelope should be as tight as possible
  + Due to symmetry we will pick the me we use the max of the $f^*$ function as the mean parameter

Bel

```{r, echo = FALSE}
mu_opt <- optimize(function(y) -f_star(y), interval = c(0.2, 0.3))$minimum
mu_opt
```


---
### Gaussian envelope

Note that

$$ 1 = \int f(x) dx = c \int f^*(x) dx \Leftrightarrow \frac{1}{\int f(x) dx} = c$$
We can thus find the proportionality constant by integrating the function $f^*$.
```{r, echo = FALSE}
c <- 1 / integrate(f_star_vec, lower = 0, upper = Inf, abs.tol = 10^(-60))$value
c %>% kable()
```

